A Wait-free Queue with Poly-logarithmic Step Complexity

### 2022-12-15 podc_paper/introduction.tex
working on related work section

A great deal of research in the past 25 years has been devoted to the design of linearizable, lock-free queues.
As well as being a fundamental data structure, queues are used in
significant concurrent applications, including OS kernels \cite{MP91}, memory reclamation \cite{?},
and sharing resources or tasks.
The lock-free MS-queue of Michael and Scott \cite{MS98} is a classic shared queue implementation.
Their design uses a singly-linked list with pointers to the front and back nodes.
To dequeue or enqueue an element, the front or back pointer is updated by a 
compare-and-swap (CAS) instruction.
If this CAS fails, the operation must retry.
In the worst case, this means that each successful CAS may cause all other processes to
fail and retry, leading to an amortized step complexity of $\Omega(p)$ per operation.
(To measure amortized step complexity of a lock-free implementation, we consider all possible finite executions
and divide the number of steps in the execution by the number of operations initated in the execution.)
Numerous papers have suggested modifications to the MS-queue \cite{list-some}, but 
all still have $\Omega(p)$ amortized step complexity as a result of
contention on the front and back of the queue.
Morrison and Afek \cite{DBLP:conf/ppopp/MorrisonA13} dubbed this the CAS retry problem.
The same problem occurs in array-based implementations of queues \cite{list-some}.

>>>>

A great deal of research in the past 25 years has been devoted to the design of linearizable, lock-free queues.
As well as being a fundamental data structure, queues are used in
significant concurrent applications, including OS kernels \cite{MP91}, memory reclamation \cite{?},
packet processing \cite{DPDK},
and sharing resources or tasks.
The lock-free MS-queue of Michael and Scott \cite{MS98} is a classic shared queue implementation.
Their design uses a singly-linked list with pointers to the front and back nodes.
To dequeue or enqueue an element, the front or back pointer is updated by a 
compare-and-swap (CAS) instruction.
If this CAS fails, the operation must retry.
In the worst case, this means that each successful CAS may cause all other processes to
fail and retry, leading to an amortized step complexity of $\Omega(p)$ per operation.
(To measure amortized step complexity of a lock-free implementation, we consider all possible finite executions
and divide the number of steps in the execution by the number of operations initated in the execution.)
Numerous papers have suggested modifications to the MS-queue \cite{DBLP:conf/opodis/HoffmanSS07,DBLP:conf/podc/KoganH14,DBLP:conf/ppopp/KoganP11,DBLP:journals/dc/Ladan-MozesS08,MKLLP22,DBLP:conf/spaa/MoirNSS05,RC17}, but 
all still have $\Omega(p)$ amortized step complexity as a result of
contention on the front and back of the queue.
Morrison and Afek \cite{DBLP:conf/ppopp/MorrisonA13} dubbed this the CAS retry problem.
The same problem occurs in array-based implementations of queues \cite{list-some}.

### 2022-12-15 podc_paper/related.tex
working on related work section

Michael and Scott \cite{DBLP:conf/podc/MichaelS96} describe the early history of concurrent queue algorithms
in the paper that described their MS-queue \cite{DBLP:conf/podc/MichaelS96}, which
is a lock-free queue that has stood the test of time.
A version of it is
included in the standard Java Concurrency Package.  %java.util.concurrent.ConcurrentLinkedQueue
However, as mentioned above, it suffers from the CAS retry problem
which means it is lock-free but not wait-free and has an amortized step complexity
of $\Theta(p)$ per operation.

A number of papers have described ways to cope with the high contention at the front
and back of the MS-queue.
Moir, Nussbaum, and Shalev~\cite{DBLP:conf/spaa/MoirNSS05} 
added an elimination array that allows an enqueue to pass its value directly
to a concurrent dequeue when the queue is empty or when concurrent operations 
can be linearized to empty the queue.
However, when there are $p$ concurrent enqueues (and no dequeues), the CAS retry problem
is still present.
The baskets queue of
Hoffman, Shalev, and Shavit~\cite{DBLP:conf/opodis/HoffmanSS07} 
is another approach to avoid contention:  concurrent enqueues are grouped into baskets
and operations within a basket can be ordered arbitrarily.
However, if there are $p$ concurrent enqueues in the same basket
the CAS retry problem occurs when they must be ordered into a list using CAS instructions.
Both modifications thus still have $\Omega(p)$ amortized step complexity.

@@

Ladan-Mozes and Shavit~\cite{DBLP:journals/dc/Ladan-MozesS08}
presented an optimistic approach to implement a queue. MS-queue uses
two \nf{CAS}es to do an enqueue: one to change the tail to the new
node and another one to change the next pointer of the previous node
to the new node. They use a doubly-linked list to do fewer
\texttt{CAS} operations in an \nf{Enqueue} than MS-queue. As in
previous algorithms, the worst case happens in the case where the
contention is high: when $p$ concurrent enqueues happen, their nodes
have to be appended to the linked list one by one. The amortized
complexity is still $\Omega(p)$ \texttt{CAS}es.

Hendler et al.~\cite{DBLP:conf/spaa/HendlerIST10} proposed a new
paradigm called flat combining. The key idea behind flat combining is
to allow a combiner who has acquired the global lock on the data
structure to learn about the requests of threads on the queue, combine
them and apply the combined results on the data structure. Their queue
is linearizable but not lock-free and they present experiments that
show their algorithm performs well in some situations.

@@

Kogan and Petrank~\cite{DBLP:conf/ppopp/KoganP11} introduced wait-free
queues based on the MS-queue and use Herlihy's helping
technique~\cite{10.1145/114005.102808} to achieve wait-freedom. Their
amortized step complexity is $\Omega(p)$ because of the helping
mechanism.

Milman et al.~\cite{DBLP:conf/spaa/MilmanKLLP18} designed a lock-free
queue supporting futures. In their queue, operations  return future
objects instead of responses. Later when the response is needed, it
can be evaluated from the future object. They also define a weaker
linearizability condition such that each operation can be linearized
between its invocation and when its future is evaluated. Their idea of
batching allows a sequence of operations to be submitted as a batch
for later execution on the MS-queue. They use some properties of the
queue size before and after a batch, similar to a part of our
work. Their queue is not wait-free: in fact, if the batch sizes are 1,
then the queue is like MS-queue.

@@

\here{Check:
Turek Shasha Prakash locking without blocking (PODS92).
Prakash Lee Johnson non blocking algorithms for concurrent data structures Tech rept 91-002 U of Florida 1991.
}

@@

\here{new result I found:}
Khanchandani and Wattenhofer \cite{KW18} gave a wait-free queue implementation
with $O(\sqrt{p})$ complexity using non-standard synchronization primitives
called half-increment and half-max, which can be viewed as particular kinds of
double-word read-modify-write operations.
They use this as evidence that their primitives can be more efficient than CAS
since previous CAS-based queues all required $\Omega(p)$ step complexity.
Our new implementation counters this argument.

>>>>

\paragraph{List-based Queues}
The MS-queue \cite{MS98} is a lock-free queue that has stood the test of time.
A version of it is
included in the standard Java Concurrency Package.  %java.util.concurrent.ConcurrentLinkedQueue
See the paper that introduced it for a survey of the early history of concurrent queues.
As mentioned above, the MS-queue suffers from the CAS retry problem because of contention at the front and back of the queue.
Thus, it is lock-free but not wait-free and has an amortized step complexity
of $\Theta(p)$ per operation.

A number of papers have described ways to reduce contention in the MS-queue.
Moir et al.~\cite{DBLP:conf/spaa/MoirNSS05} 
added an elimination array that allows an enqueue to pass its enqueued value directly
to a concurrent dequeue when the queue is empty or when concurrent operations 
can be linearized to empty the queue.
However, when there are $p$ concurrent enqueues (and no dequeues), the CAS retry problem
is still present.
The baskets queue of
Hoffman, Shalev, and Shavit~\cite{DBLP:conf/opodis/HoffmanSS07} 
attempts to reduce contention by grouping concurrent enqueues into baskets.
An enqueue that fails its CAS is put in the basket with the eqnueue that succeeded.
Enqueues within a basket determine their order among themselves without having to access the back of the queue.
However, if $p$ concurrent enqueues are in the same basket
the CAS retry problem occurs when they order themselves using CAS instructions.
Both modifications still have $\Omega(p)$ amortized step complexity.

Kogan and Herlihy \cite{DBLP:conf/podc/KoganH14} described how to use futures to improve
the performance of the MS-queue.
Operations return future
objects instead of responses. Later, when an operation's response is needed, it
can be evaluated using the corresponding future object.
This allows batches of enqueues or dequeues to be done at once on an MS-queue.
However, the implementation satisfies a weaker correctness condition than linearizability.
Milman-Sela et al.~\cite{MKLLP22} extended this approach to allow batches
to mix enqueues and dequeues.
\here{Is this worth saying:
They use some properties of the queue size before and after a batch, similar to a part of our work.}
In the worst case, where operations require their response right away,
batches have size 1, and both of these implementations behave like a standard MS-queue.

Ladan-Mozes and Shavit~\cite{DBLP:journals/dc/Ladan-MozesS08}
presented an optimistic  queue implementation. 
%MS-queue uses
%two \nf{CAS}es to do an enqueue: one to change the tail to the new
%node and another one to change the next pointer of the previous node
%to the new node. 
In the MS-queue, an enqueue requires two CAS steps.
The optimistic queue uses a doubly-linked list to reduce the number of
\texttt{CAS} instructions to one in the best case. 
Pointers in the doubly-linked list can be inconsistent, but are fixed when needed by traversing the list.
Although this fixing is rare in practice, it yields an amortized complexity of $\Omega(qp)$ 
steps per operation for worst-case executions.

Kogan and Petrank~\cite{DBLP:conf/ppopp/KoganP11} 
added Herlihy's helping
technique~\cite{10.1145/114005.102808} to the MS-queue to obtain
a wait-free queue.
Ramalhete and Correia \cite{RC17} added a different helping mechanism.
In both cases, the helping mechanisms only add to the step complexity of the MS-queue.

@@

% OMIT, since it is not lock-free:
%Hendler et al.~\cite{DBLP:conf/spaa/HendlerIST10} proposed a new
%paradigm called flat combining. The key idea behind flat combining is
%to allow a combiner who has acquired the global lock on the data
%structure to learn about the requests of threads on the queue, combine
%them and apply the combined results on the data structure. Their queue
%is linearizable but not lock-free and they present experiments that
%show their algorithm performs well in some situations.

@@

@@

\here{Check:
John Giacomoni et al. FastForward
https://dl-acm-org.ezproxy.library.yorku.ca/doi/10.1145/1345206.1345215
}

@@

\here{new result I found:}
Khanchandani and Wattenhofer \cite{KW18} gave a wait-free queue implementation
with $O(\sqrt{p})$ step complexity using non-standard synchronization primitives
called half-increment and half-max, which can be viewed as particular kinds of
double-word read-modify-write operations.
They use this as evidence that their primitives can be more efficient than CAS
since previous CAS-based queues all required $\Omega(p)$ step complexity.
Our new implementation counters this argument.
